The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-09 paper/ms.tex
minor wording changes, plus additions to the intro of s4 that will hopefully help put context of the different sims up front and satisfy some of Chervin's comments

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a revolution as a
result of \textit{Gaia} Data Release 2 (DR2).  Recent major discoveries
include the remnants of a major merger \citep{2018ApJ...860L..11K,
2018Natur.563...85H, 2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space
``spiral'' in the solar neighborhood \citep{2018Natur.561..360A} possibly
indicating local substructure infall \citep{2018MNRAS.481.1501B,
2018arXiv180800451L}, and a gap
suggestive of perturbation by a dark matter subtructure in the tidal stream
GD1 \citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

@@

With the advent of 6D phase-space measurements over a relatively large
($\gtrsim 2$ kpc) volume from the \textit{Gaia} satellite, the study of stars'
actions has gained new popularity. One reason is dimensionality reduction---an
individual stellar orbit is concisely described by three actions, as opposed
to six phase space coordinates. Second, under the assumption of a phase-mixed
system, the dynamical properties of a population of stars should be uniquely a
function of their actions and independent of the conjugate angles, allowing
investigation of the relationship between \emph{orbital} properties of stars
and other intrinsic, and at least partially invariant, properties such as age
or metallicity \citep{2018ApJ...867...31B, 2018arXiv180803278T,
2018MNRAS.481.4093S, 2019arXiv190304030G, 2019arXiv190309320D,
2019MNRAS.486.1167B}. Actions also provide a convenient basis for constructing
models of the stellar distribution function
\citep[e.g.][]{1915MNRAS..76...70J, 1985ApJ...295..388V, 2017ApJ...839...61T},
or for associating stars with similar dynamical properties, e.g., to
potentially determine membership in moving groups.

@@

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. However, the Galactic potential is not strictly
axisymmetric. This has been known for some time \citep[e.g.][]{1957AJ.....62...93K,2009MNRAS.396L..56M, 2012ApJ...750L..41W}, but with the vast improvement in the quality of phase-space
measurements due to \textit{Gaia} this assumption is increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A, 2019MNRAS.485.3134L}. Even if this assumption were
close enough for many purposes, the parameters used in axisymmetric models of the Galactic potential may
be inadequately constrained by current observations.

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' and ``global'' Galactic
midplanes are not the same. For stars relatively far from the Sun, their local midplane
will not agree with our local midplane extrapolated onto their position due to
relatively small-scale variations in the density of the gas and stellar
distribution. Converting positions and velocities of more distant stars from a
heliocentric to a Galactocentric coordinate system thus introduces a systematic
bias in the $z$ coordinate, and consequently the actions computed for these stars.
The further from the solar neighborhood the target star, the more likely the
mismatch will result in large systematic uncertainty, especially in the
vertical action $J_z$. A similar argument applies to any remaining uncertainty
in measurements of the Galactocentric radius of the Sun and to variations in
the LSR.

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/cartoon.pdf}
\end{center}
\caption{Cartoon approximation showing the effect an error in the
determination of the coordinate midplane can have on orbit integration and
action estimation. The $x$-axis shows the orbital phase and the $y$-axis the
vertical height. The top gray curve depicts an example ``true'' orbit
oscillating about the true midplane (horizontal solid gray line). Consider an
observer who erroneously assumes the midplane is located at the horizontal
dashed line. Suppose this observer measures the phase-space position of the
star at two different orbital phases (teal and orange points). If the observer
were to then integrate the star's orbit using a model potential with the
erroneous midplane, they would obtain the teal and orange curves for the
star's orbit, respectively. The actions estimated from these two erroneous
measurements would subsequently differ, both from each other and from the true measurement (in the potential with the correct coordinate system).
Hence an incorrect midplane in the potential model assumed will induce
phase-dependence in the actions estimated for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}

@@

\subsection{Effect of Midplane Offset} \label{ssec:cartoon}
We present a cartoon approximation in Figure~\ref{fig:cartoon} to show how an
inaccurate or erroneous determination of the midplane leads to a phase-dependence of the actions
calculated from an orbit specified by a phase-space point and an assumed
potential model. The $y$-axis corresponds to the vertical height of the orbit
and the $x$-axis shows orbital phase. The solid gray curve indicates the
``true'' orbit of the star; that is, its orbit as it oscillates around the
true midplane. The dashed gray line is the midplane location in the potential
model used by an observer to integrate the orbit of the star and hence
estimate its actions, which is offset from the true midplane of the star's
orbit. The model potential is otherwise identical to the one in which the star
is actually moving.

Now suppose this observer makes a measurement of the star's position and
velocity at the teal point or the orange point (i.e. at two different orbital
phases). Then the teal and orange curves correspond to the orbits that the
observer would compute for each point based on the potential model with the
offset midplane. In action space, this would correspond to a different value
of $J_z$ for the teal and orange points. In this way, assuming the wrong
coordinate system induces a phase-dependence in the actions estimated for the
star, which in the correct potential (in this example, the one with the
correct midplane) should be phase-independent.

@@

Before turning to numerical methods, we first write down analytically the
systematic error in the actions induced from either a midplane offset or a
Galactocentric radius offset. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of thin disk-like orbits and a good description of thick disk-like orbits in
an axisymmetric potential, i.e., ignoring the influence of the Galactic bar and
spiral arms. We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle
\citep[e.g.][]{2017MNRAS.465...76M}.

Under this approximation, we can write down the cylindrical components
of the orbits as
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\kappa t + \alpha)} \\
z(t) &= A_z \sin{(\nu t + \beta)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the radial and vertical frequencies, $\Omega_c
\equiv v_c/R_g$ is the rotation frequency, $A_z$ and $A_R$ are the amplitudes of
the stellar motion in the vertical and radial coordinates, and $\alpha$ and
$\beta$ are arbitrary constants. Similarly, the velocities of the orbit are
given by:
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= \kappa A_R \cos{(\kappa t + \alpha)} \\
v_z(t) &= \nu A_z \cos{(\nu t + \beta)}
\text{.}
\end{split}
\eeq

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the observed values in $J_z$ for the thick-disk (top
panel) and thin-disk (bottom panel) orbits assuming a $z$ offset of
$100\,\pc$. One can see that if the observed $z$ values have a bias (from e.g.
an incorrectly computed midplane), then the induced error distribution in
$J_z$ is decidedly non-Gaussian. Therefore, any sort of error propagation must
take this into account. Intuition for the shape of each panel is given in the
text. We also plot one half the 95\uth{} percentile minus the 5\uth{}
percentile of each distribution as a horizontal arrow anchored on the true
$J_z$ value. We call this $\Delta J_z$ and will use it (along with the
similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to empirically describe
the error distribution. We see that $\Delta J_z$ roughly corresponds to the
distance from the true $J_z$ value to one of the modes of the distribution of
observed $J_z$ values. Similar plots for $J_R$ induced by a $z$ offset and
$J_R$ and $J_{\phi}$ induced by an $x$ offset are given in
Appendix~\ref{app:hist}.}
\label{fig:Jz_hist}
\end{figure}

We quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the \emph{observed} actions. We will consider
three orbits in the model potential described in
Section~\ref{ssec:action_comp} that are typical of stars in the thin
disk, thick disk, and halo. We summarize their initial
positions in phase space and the actions computed by integrating their orbits
in the correct potential in Table~\ref{tab:orbits}. Each orbit, integrated
without systematic coordinate errors, is plotted in Appendix~\ref{app:orbits}.
We will refer to these orbits by their names (thin-disk, thick-disk, halo)
henceforth.

We begin with the thick-disk orbit. Consider an immortal observer who can
measure the thick-disk star's phase-space position at many different times
(and hence different orbital phases). However, the immortal observer uses a
coordinate system in which the midplane is systematically offset in height by
$100\,\pc$ from its actual location --- i.e. we subtract the vector $(0, 0,
100)\,\pc$ from each position in the orbit. This corresponds to the immortal
observer physically located at e.g. the position $(8, 0, 0)\,\kpc$ in the
coordinate system of the true potential, but erroneously thinking they are
located at $(8, 0, 0.1)\,\kpc$.

We consider that the observer makes a measurement every $\Myr$. At
each measurement, we compute the actions as outlined above using the true
potential but by specifiying the star's starting position using the
systematically offset coordinate system. Essentially we are shifting the
integrated orbit and then reintegrating at each point along the shifted orbit,
where the shifted orbit is the observed orbit. The erroneous computed actions
for each phase-space starting point are shown for the first $\Gyr$ of the
orbit in the upper panels
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $5\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the observer this would
appear as time-dependence in the actions, oscillating around
their value using the correct coordinate system. This time dependence comes even though
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality, we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $R$ or $z$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}).  In the following we
therefore quote percentile ranges for the possible values computed for each
action as a proxy for the effect of these systematic errors in the coordinate
system.

For a systematic offset in $z$ (upper panels), the 95\textsuperscript{th}
minus 5\textsuperscript{th} percentiles are $2.2$ and $6.2\,\actunit$ for
$J_R$ and $J_z$, respectively. These are fractional errors of $5.7\,\%$ and
$86\,\%$, respectively. The error induced in $J_{\phi}$ is negligible, as
expected since $J_{\phi}$ only depends on the $x$- and $y$-components of the
position and velocity of the stars.\footnote{In practice, however, $J_{\phi}$
is computed as part of the torus-fitting method.} It is worth pointing out
that a $100\,\pc$ error in an orbit with $z_{\text{max}}=850\,\pc$ --- a
$12\%$ error --- induced an $86\%$ error in the computation of $J_z$.

For a systematic offset in $x$ (or distance to the Galactic center), the
95\textsuperscript{th} minus 5\textsuperscript{th} percentiles are $6.9$,
$47$, and $0.71\,\actunit$ for $J_R$, $J_{\phi}$ and $J_z$, respectively.
These are fractional errors of $21\%$, $3.1\%$ and $3.1\%$, respectively ---
despite only a $1.2\%$ error in the distance to the Galactic center.

In Figure~\ref{fig:Jz_hist} we plot a histogram of the observed values of
$J_z$ for the thick-disk orbit (top panel) and thin-disk orbit (bottom panel)
assuming a $z$ offset of $100\,\pc$ (i.e. as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}). The true value of $J_z$ is plotted as a
vertical dashed line. Here we see that the errors in $J_z$ induced by a
systematic offset in $z$ are non-Gaussian and bimodal. Furthermore, neither of
the modes lie at zero. In other words, neither of the modes of the
distribution of observed $J_z$ values are centered on the true $J_z$ (i.e. the
$J_z$ computed with no offset in $z$). In the case of the thin-disk orbit
(bottom panel), we see that, in addition to the prior complications, the error
distribution is not even centered on null. This comes about when the midplane
error is $\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit, equivalent to $A_z$ in the epicyclic approximation (see
Section~\ref{ssec:epi_action}).

In Appendix~\ref{app:hist} we plot the same histogram as in
Figure~\ref{fig:Jz_hist}, but for the observed distributions of $J_R$ induced
by a $z$ offset (upper left panel of Figure~\ref{fig:many_orbit_wrong_ref})
and the observed distributions of $J_R$ and $J_{\phi}$ errors induced by an
$x$ offset (lower left and lower center panels of
Figure~\ref{fig:many_orbit_wrong_ref}, respectively). We find similar error
distributions as in Figure~\ref{fig:Jz_hist}, with the exception that the
observed $J_R$ distribution induced by an $x$ offset more closely resembles a
Gaussian.

@@

We now describe the intuition behind the shape of Figure~\ref{fig:Jz_hist}.
Consider first the thick-disk orbit (top panel), where the offset in $z$ is
much less than $z_{\text{max}}$ of the orbit. The peaks in the distribution
correspond to the turning points of the orbit (or points of maximum vertical
amplitude), where $v_z \sim 0$ and where the star is on most of it orbit. This
is why the distribution is peaked on these values. For the thin-disk orbit
(bottom panel), the offset in $z$ is comparable to $z_{\text{max}}$. Now,
there will be some points in the orbit where $v_z = 0$ and $z=0$ (in the
observed, erroneous coordinate system). At these points, the computed $J_z$
will vanish. The asymmetry and systematic offset then comes about because of
the constraint that $J_z \geq 0$.\footnote{This is similar to arguments in
cosmology for why gravity produces non-Gaussianity in the density field, since
the density cannot become negative but it can grow arbitrarily large.}

Empirically describing the error distribution shown in
Figure~\ref{fig:Jz_hist} is difficult. Of course, Gaussian summary statistics
are not adequate. We therefore elect to measure this error by computing one
half the 95\uth{} percentile minus the 5\uth{} percentile of the distribution
of observed action values. We plot this quantity, which we refer to as $\Delta J_i$ for each action, in Figure~\ref{fig:Jz_hist} as a
horizontal arrow anchored on the true action value. Because of the bimodality
of the error distribution, this quantity roughly measures the distance from
the true action value to the peak of one of the modes. Furthermore, this
bimodality also implies that $\Delta J_i$ is not very sensitive to the exact
percentiles used. This summary statistic does not reflect the bias induced
when the midplane error is $\gtrsim z_{\text{max}}$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(-190, 0, v_z)\,\kms$, where we vary $v_z$. We show
this for a $z$ offset of $10, 50, 100\,\pc$ (orange, teal, and green
respectively). As before, the error ($\Delta J_z$) is one half the 95\uth{}
minus 5\uth{} percentile of the distribution of observed $J_z$ values over the
course of the orbit. There are large errors for thin-disk like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

@@

\section{Coordinate System Offsets II. Azimuthal Midplane Variations in
Simulations}
\label{sec:local_fire}

We now consider a second source of systematic offsets in the Galactocentric
coordinate system: intrinsic variations of the midplane across the disk. In
this work we specifically consider azimuthal variations in the midplane as
defined by the stellar mass density. We use two sets of simulations of Milky
Way-mass galaxies to estimate the size of this effect.

>>>>

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a revolution as a
result of \textit{Gaia} Data Release 2 (DR2).  Recent major discoveries
include the remnants of a major merger \citep{2018ApJ...860L..11K,
2018Natur.563...85H, 2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space
``spiral'' in the solar neighborhood \citep{2018Natur.561..360A} possibly
indicating either local substructure infall \citep{2018MNRAS.481.1501B,
2018arXiv180800451L} or bar buckling \citep{2018arXiv181109205K}, and a gap
suggestive of perturbation by a dark matter subtructure in the tidal stream
GD1 \citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

@@

With the advent of 6D phase-space measurements over a relatively large
($\gtrsim 2$ kpc) volume from the \textit{Gaia} satellite, the study of stars'
actions has gained new popularity. One reason is dimensionality reduction---an
individual stellar orbit is concisely described by three actions, as opposed
to six phase space coordinates. Second, under the assumption of a phase-mixed
system, the dynamical properties of a population of stars should be uniquely a
function of their actions and independent of the conjugate angles, allowing
investigation of the relationship between \emph{orbital} properties of stars
and other intrinsic, and at least partially invariant, properties such as age
or metallicity \citep{2018ApJ...867...31B, 2018arXiv180803278T,
2018MNRAS.481.4093S, 2019arXiv190304030G, 2019arXiv190309320D,
2019MNRAS.486.1167B}. Actions also provide a convenient basis for constructing
models of the stellar distribution function
\citep[e.g.][]{1915MNRAS..76...70J, 1985ApJ...295..388V, 2017ApJ...839...61T},
or for associating stars with similar dynamical properties (e.g. to determine
membership in moving groups).

@@

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. However, the Galactic potential is not strictly
axisymmetric; with the vast improvement in the quality of phase-space
measurements due to \textit{Gaia}, this assumption appears increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this assumption were
close enough for many purposes, the parameters used in axisymmetric models may
be inadequately constrained by current observations.

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' Galactic midplane is not the same as the ``global'' Galactic
midplane. For stars relatively far from the Sun, their local midplane location
will not agree with our local midplane extrapolated onto their position due to
relatively small-scale variations in the density of the gas and stellar
distribution. Converting positions and velocities of more distant stars from a
heliocentric to a Galactocentric coordinate system introduces a systematic
bias in the $z$ coordinate, and hence the actions computed for these stars.
The further from the solar neighborhood the target star, the more likely the
mismatch will result in large systematic uncertainty, especially in the
vertical action $J_z$. A similar argument applies to any remaining uncertainty
in measurements of the Galactocentric radius of the Sun and to variations in
the LSR.

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/cartoon.pdf}
\end{center}
\caption{Cartoon approximation showing the effect an error in the
determination of the coordinate midplane can have on orbit integration and
action estimation. The $x$-axis shows the orbital phase and the $y$-axis the
vertical height. The top gray curve depicts an example ``true'' orbit
oscillating about the true midplane (horizontal solid gray line). Consider an
observer who erroneously assumes the midplane is located at the horizontal
dashed line. Suppose this observer measures the phase-space position of the
star at two different orbital phases (teal and orange points). If the observer
were then to integrate the star's orbit using a model potential with the
erroneous midplane, they would obtain the teal and orange curves for the
star's orbit, respectively. The actions estimated from these two  erroneous
orbits would subsequently differ, both from each other and from the actions
estimated for the true orbit (in the potential with the correct midplane).
Hence an incorrect midplane in the potential model assumed will induce
phase-dependence in the actions estimated for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}

@@

\subsection{Effect of Midplane Offset} \label{ssec:cartoon}
We present a cartoon approximation in Figure~\ref{fig:cartoon} to show how an
inaccurate or erroneous determination of the midplane (and hence the
zero-point of the $z$ coordinate) leads to a phase-dependence of the actions
calculated from an orbit specified by a phase-space point and an assumed
potential model. The $y$-axis corresponds to the vertical height of the orbit
and the $x$-axis shows orbital phase. The solid gray curve indicates the
``true'' orbit of the star; that is, its orbit as it oscillates around the
true midplane. The dashed gray line is the midplane location in the potential
model used by an observer to integrate the orbit of the star and hence
estimate its actions, which is offset from the true midplane of the star's
orbit. The model potential is otherwise identical to the one in which the star
is actually moving.

Now suppose this observer makes a measurement of the star's position and
velocity at the teal point or the orange point (i.e. at two different orbital
phases). Then the teal and orange curves correspond to the orbits that the
observer would compute for each point based on the potential model with the
offset midplane. In action space, this would correspond to a different value
of $J_z$ for the teal and orange points. In this way, assuming the wrong
coordinate system induces a phase-dependence in the actions estimated for the
star, which in the correct potential (in this example, the one with the
correct midplane) should be phase-independent. This example uses an offset in $z$, but one would expect an analogous effect from an offset in another coordinate, such as the distance to the Galactic center or the local standard of rest.

@@

Before turning to numerical methods, we derive analytic expressions for the
systematic error in the actions induced from offsets in the position (e.g. the midplane or
Galactic center distance) or velocity (e.g. the local standard of rest) of the zero-point in the assumed coordinate system. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of thin disk-like orbits and a good description of thick disk-like orbits in
an axisymmetric potential (i.e. ignoring the influence of the Galactic bar and
spiral arms). We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle
\citep[e.g.][]{2017MNRAS.465...76M}.

Under this approximation, we can write down the cylindrical components
of the orbits as
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\kappa t + \alpha)} \\
z(t) &= A_z \sin{(\nu t + \beta)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the radial and vertical frequencies, $\Omega_c
\equiv v_c/R_g$ is the orbital frequency of the guiding center, $A_z$ and $A_R$ are the amplitudes of
the stellar motion in the vertical and radial coordinates, and $\alpha$ and
$\beta$ are arbitrary phases. Similarly, the velocities of the orbit are
given by:
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= \kappa A_R \cos{(\kappa t + \alpha)} \\
v_z(t) &= \nu A_z \cos{(\nu t + \beta)}
\text{.}
\end{split}
\eeq

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the computed values of $J_z$ for the thick-disk (top
panel) and thin-disk (bottom panel) orbits assuming a $z$ offset of
$100\,\pc$, at different orbital phases. One can see that if the observed $z$ values have a bias (from e.g.
an incorrectly computed midplane), then the induced error distribution in
$J_z$ is decidedly non-Gaussian. Therefore, any sort of error propagation must
take this into account. Intuition for the shape of each panel is given in the
text. We also plot one half the 95\uth{} percentile minus the 5\uth{}
percentile of each distribution as a horizontal arrow anchored on the true
$J_z$ value. We call this $\Delta J_z$ and will use it (along with the
similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to empirically describe
the error distribution. We see that $\Delta J_z$ roughly corresponds to the
distance from the true $J_z$ value to one of the modes of the distribution of
computed $J_z$ values. Similar plots for $J_R$ induced by a $z$ offset and
$J_R$ and $J_{\phi}$ induced by an $x$ offset are given in
Appendix~\ref{app:hist}.}
\label{fig:Jz_hist}
\end{figure}

We now quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the actions calculated from the observed position and velocity of a star. We consider
three orbits in the model potential described in
Section~\ref{ssec:action_comp} that are typical of stars in the stellar thin
disk, stellar thick disk, and the stellar halo. We summarize their initial
positions in phase space and the actions computed by integrating their orbits
in the correct potential in Table~\ref{tab:orbits}. Each orbit, integrated
without systematic coordinate errors, is plotted in Appendix~\ref{app:orbits}.
We will refer to these orbits by their names (thin-disk, thick-disk, halo)
henceforth.

We begin with the thick-disk orbit. Consider an observer who can
measure the thick-disk star's phase-space position at many different times
(and hence different orbital phases), but does so using a
coordinate system in which the midplane is systematically offset in height by
$100\,\pc$ from its actual location. To model this we subtract the vector $(0, 0,
100)\,\pc$ from each position in the orbit. This corresponds to an
observer physically located at e.g. the position $(8, 0, 0)\,\kpc$ in the
coordinate system of the true potential, but erroneously thinking they are
located at $(8, 0, 0.1)\,\kpc$.

We consider that the observer makes a measurement every $\Myr$. At
each measurement, we compute the actions as outlined above using the true
potential but by specifiying the star's starting position using the
systematically offset coordinate system. Essentially we are shifting the
integrated orbit and then reintegrating at each point along the shifted orbit. The actions computed using the offset coordinate system
for each phase-space starting point are shown for the first $\Gyr$ of the
orbit in the upper panels
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $5\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the observer this would
appear as time-dependence in the actions, which appear to oscillate around
their value using the correct coordinate system even though in this example
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $R$ or $z$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}).  In the following we
therefore quote percentile ranges for the possible values computed for each
action as a proxy for the effect of these systematic errors in the coordinate
system.

For a systematic offset in $z$ (upper panels), the 95\textsuperscript{th}
minus 5\textsuperscript{th} percentiles are $2.2$ and $6.2\,\actunit$ for
$J_R$ and $J_z$, respectively. These are fractional errors of $5.7\,\%$ and
$86\,\%$, respectively. The error induced in $J_{\phi}$ is negligible, as
expected since $J_{\phi}$ only depends on the $x$- and $y$-components of the
position and velocity of the stars.\footnote{In practice, however, $J_{\phi}$
is computed as part of the torus-fitting method.} It is worth pointing out
that a $100\,\pc$ error in an orbit with $z_{\text{max}}=850\,\pc$ --- a
$12\%$ error---induced an $86\%$ error in the computation of $J_z$.

For a systematic offset in $x$ (or distance to the Galactic center), the
95\textsuperscript{th} minus 5\textsuperscript{th} percentiles are $6.9$,
$47$, and $0.71\,\actunit$ for $J_R$, $J_{\phi}$ and $J_z$, respectively.
These are fractional errors of $21\%$, $3.1\%$ and $3.1\%$ in these actions, respectively, despite only a $1.2\%$ error in the distance to the Galactic center.

In Figure~\ref{fig:Jz_hist} we plot a histogram of the values of
$J_z$ computed for the thick-disk orbit (top panel) and thin-disk orbit (bottom panel)
assuming a $z$ offset of $100\,\pc$ (i.e. as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}), as a function of orbital phase. The true value of $J_z$ is plotted as a
vertical dashed line. The systematic error in $J_z$ induced by a
systematic offset in $z$ is non-Gaussian and bimodal; neither of the modes is centered on the true $J_z$ (i.e. the $J_z$ computed with no offset in $z$). In the case of the thin-disk orbit
(bottom panel), we see that, in addition to the prior complications, the
distribution is not even centered on the true value. This comes about when the midplane
error is $\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit, equivalent to $A_z$ in the epicyclic approximation (see
Section~\ref{ssec:epi_action}).

In Appendix~\ref{app:hist} we plot the same histogram as in
Figure~\ref{fig:Jz_hist}, but for the distributions of $J_R$ induced
by a $z$ offset (upper left panel of Figure~\ref{fig:many_orbit_wrong_ref})
and the distributions of $J_R$ and $J_{\phi}$ errors induced by an
$x$ offset (lower left and lower center panels of
Figure~\ref{fig:many_orbit_wrong_ref}, respectively). We find similar error
distributions as in Figure~\ref{fig:Jz_hist}, with the exception that the
observed $J_R$ distribution induced by an $x$ offset more closely resembles a
Gaussian.

@@

We now describe the intuition behind the shape of Figure~\ref{fig:Jz_hist}.
Consider first the thick-disk orbit (top panel), where the offset in $z$ is
much less than $z_{\text{max}}$ of the orbit. The peaks in the distribution
correspond to the turning points of the orbit (or points of maximum vertical
amplitude), where $v_z \sim 0$ and where the star is on most of it orbit. This
is why the distribution, which is calculated at evenly spaced time intervals, peaks at these values. For the thin-disk orbit
(bottom panel), the offset in $z$ is comparable to $z_{\text{max}}$. Now,
there will be some points in the orbit where $v_z = 0$ and $z=0$ (in the
erroneous coordinate system). At these points, the computed $J_z$
will vanish. The asymmetry and systematic offset then comes about because of
the constraint that $J_z \geq 0$.\footnote{This is similar to arguments in
cosmology for why gravity produces non-Gaussianity in the density field, since
the density cannot become negative but it can grow arbitrarily large.}

Gaussian summary statistics are clearly insufficient to describe the distribution shown in
Figure~\ref{fig:Jz_hist}. We therefore elect to measure this error by computing one
half the 95\uth{} percentile minus the 5\uth{} percentile of the distribution
of action values. We refer to this quantity as $\Delta J_i$ for each
action value. We plot this quantity in Figure~\ref{fig:Jz_hist} as a
horizontal arrow anchored on the true action value. Because of the bimodality
of the error distribution, this quantity roughly measures the distance from
the true action value to the peak of one of the modes. Furthermore, this
bimodality also implies that $\Delta J_i$ is not very sensitive to the exact
percentiles used. This summary statistic does not reflect the bias induced
when the midplane error is $\gtrsim z_{\text{max}}$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(-190, 0, v_z)\,\kms$, where we vary $v_z$. We show
this for a $z$ offset of $10, 50, 100\,\pc$ (orange, teal, and green
respectively). As before, the error ($\Delta J_z$) is one half the 95\uth{}
minus 5\uth{} percentile of the distribution of $J_z$ values over the
course of the orbit. There are large errors for thin-disk like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

@@

\section{Coordinate System Offsets from Azimuthal Midplane Variations}
\label{sec:local_fire}
